A Statistical Dynamical Model to Predict Extreme Events and Anomalous Features in Shallow Water Waves with Abrupt Depth Change

### 2018-11-29 paper1_tKdV/tKdV_theory.tex
DQ_final

Controlled laboratory experiments were carried out in  \cite{bolles2018anomalous} to examine the statistical behavior of surface waves crossing an ADC. In these experiments, nearly unidirectional waves are generated by a paddle wheel and propagate through a long, narrow wave tank. Midway through, the waves encounter a step in the bottom topography, and thus abruptly transition to a shallower depth. The paddle wheel is forced with a pseudo-random signal intended to mimic a Gaussian random sea upstream of the ADC. In particular, the paddle angle is specified as
\[
\theta\left(t\right)=\theta_{0}+\Delta\theta\sum_{n=1}^{N}a_{n}\cos\left(\omega_{n}t+\delta_{n}\right),\;E\sim\left(\Delta\theta\right)^{2}\sum_{n}a_{n}^{2}\omega_{n}^{2}.
\]
where the weights $a_n$ are Gaussian in spectral space with peak frequency $\omega_p$ and the phases $\delta_n$ are uniformly distributed random variables. The peak frequency gives rise to a characteristic wavelength $\lambda_c$ which can be estimated from the dispersion relation. The energy $E$ injected into the system is determined by the angle amplitude $\Delta\theta$, which is the main control parameter varied in \cite{bolles2018anomalous}. Optical measurements of the free surface reveal a number of surprising statistical features:
\begin{itemize}
\item Distinct statistics are found between the incoming and outgoing wave
disturbances: incoming waves display near-Gaussian statistics,
while outgoing waves skew strongly towards positive displacement.
\item The degree of non-Gaussianity present in the outgoing waves depends on the injected energy $E$: 
larger energies generate stronger skewness in the surface displacement PDFs and more extreme events.
\item \textcolor{blue} {Compared to the incoming wave train, the power spectrum of the outgoing wave field decays more slowly, which indicates that the anomalous behavior is associated with an elevated level of high frequencies.}
%{The waves also show different slopes in the power spectra at upstream and downstream locations}.
\end{itemize}

@@

The Korteweg-de Vries (KdV) equation is a one-dimensional, deterministic model capable of describing {\color{blue} (weak?)} surface wave turbulence. More specifically, KdV is leading-order approximation for surface waves governed by a balance of nonlinear and dispersive effects, valid in an appropriate far-field limit \cite{johnson1997modern}. Moreover, KdV has been adapted to describe waves propagating over variable depth  \cite{johnson1997modern}. 
Here, we consider the variable-depth KdV equation truncated at wavenumber $\Lambda$ (with $J=2\Lambda+1$ grid points) in order to generate weakly turbulent dynamics. The surface displacement is described by the state variable $u_{\Lambda}^{\pm}\left(x,t\right)$ with superscript `$-$' for the incoming waves and `$+$' for the outgoing waves.
The Galerkin truncated variable $u_{\Lambda}=\sum_{1\leq\left|k\right|\leq\Lambda}\hat{u}_{k}\left(t\right)e^{ikx}$ is normalized with zero mean $\hat{u}_{0}=0$ and unit energy $2\pi\sum_{k=1}^{\Lambda}\left|\hat{u}_{k}\right|^{2}=1$, which are conserved quantities. Here, $u_{\Lambda}\equiv\mathcal{P}_{\Lambda}u$ denotes the subspace projection. The evolution of  $u_{\Lambda}^{\pm}$ is governed by the truncated KdV equation with depth change $D_{\pm}$
\begin{equation}
\frac{\partial u_{\Lambda}^{\pm}}{\partial t}+\frac{D_{\pm}^{-3/2}}{2}E_{0}^{1/2}L_{0}^{-3/2}\frac{\partial}{\partial x}\mathcal{P}_{\Lambda}\left(u_{\Lambda}^{\pm}\right)^{2}+D_{\pm}^{1/2}L_{0}^{-3}\frac{\partial^{3}u_{\Lambda}^{\pm}}{\partial x^{3}}=0. \label{eq:tKdV_nondim}
\end{equation}
Equation [\ref{eq:tKdV_nondim}] is non-dimensionalized on the periodic
domain $x\in\left[-\pi,\pi\right]$. The depth is assumed to be unit $D_{-}=1$ before the ADC and $D_{+}<1$ after the ADC. 
The conserved Hamiltonian can be decomposed as
\begin{align*}
\mathcal{H}_{\Lambda}^{\pm} & =  D_{\pm}^{-3/2}E_{0}^{1/2}L_{0}^{-3/2}H_{3}\left(u_{\Lambda}^{\pm}\right)-D_{\pm}^{1/2}L_{0}^{-3}H_{2}\left(u_{\Lambda}^{\pm}\right),\\
& H_{3}\left(u\right) =  \frac{1}{6}\int_{-\pi}^{\pi}u^{3}dx,\;H_{2}\left(u\right)=\frac{1}{2}\int_{-\pi}^{\pi}\left(\frac{\partial u}{\partial x}\right)^{2}dx.
\end{align*}
where we refer to $H_3$ as the cubic term and $H_2$ the quadratic term.
We introduce parameters $\left(E_{0},L_{0},\Lambda\right)$
based on the following assumptions:
\begin{itemize}
\item The wavenumber truncation $\Lambda$ is fixed at a moderate value
for generating weakly turbulent dynamics.
\item The state variable $u_{\Lambda}^{\pm}$ is normalized with zero mean
and unit energy, $\mathcal{M}\left(u_{\Lambda}\right)=0,\mathcal{E}\left(u_{\Lambda}\right)=1$,
which are conserved during evolution. Meanwhile, $E_{0}$ characterizes the total
energy injected into the system based on the driving amplitude $\Delta \theta$.
\item The length scale of the system $L_{0}$ is chosen so that the resolved scale $\Delta x=2\pi L_{0}/J$ is comparable to the the characteristic wave length $\lambda_{c}$ from the experiments.
\end{itemize}
% ORIGINAL
%The intuition for the distinct model dynamics comes from the balance
%between the cubic and quadratic terms in the Hamiltonian $\mathcal{H}_{\Lambda}^{\pm}$.
%After the depth change, $D_{+}<1$, more weight is added in the cubic
%term, $H_{3}$, for stronger nonlinearity and weaker dispersion for
%the third-order derivative term reflected by the smaller coefficient
%for $H_{2}$ in the Hamiltonian. 
Some intuition for how \eqref{eq:tKdV_nondim} produces different dynamics on either side of the ADC can be gained by considering the relative contributions of $H_3$ and $H_2$ in the Hamiltonian $\mathcal{H}_{\Lambda}^{\pm}$. The depth change, $D_{+}<1$, increases the weight of $H_3$ and decreases that of $H_2$, thereby promoting the effects of nonlinearity over dispersion and creating conditions favorable for extreme events. Since $\frac{\partial u}{\partial x}$ is the slope of the wave height, $H_{2}\left(u\right)$ measures the
wave slope energy.

@@

The model parameters $\left(E_{0},L_{0},\Lambda\right)$ in [\ref{eq:tKdV_nondim}]
can be directly linked to the basic scales from the physical problem.
The important parameters that characterize the experiments \cite{bolles2018anomalous}
include: $\epsilon=\frac{a}{H_{0}}$ the wave amplitude $a$ to (upstream) water
depth $H_{0}$ ratio; $\delta=\frac{H_{0}}{\lambda_{c}}$ the water
depth to wavelength $\lambda_{c}$ ratio; and $D_{0}=\frac{d}{H_{0}}$ 
the depth ratio with upstream value $d=H_{0}$
and downstream value $d<H_{0}$. 
By comparing the characteristic physical scales, the normalized TKdV equation [\ref{eq:tKdV_nondim}] can be
linked to these experimental parameters via
\begin{equation}
L_{0}=6^{\frac{1}{3}}\left(M\epsilon^{\frac{1}{2}}\delta^{-1}\right),\;E_{0}=\frac{27}{2}\gamma^{-2}\left(M\epsilon^{\frac{1}{2}}\delta^{-1}\right),\label{eq:params}
\end{equation}
where $M$ defines the computational domain size $M\lambda_{c}$
as $M$-multiple of the characteristic wavelength $\lambda_{c}$,
and $\gamma=\frac{U}{a}$ normalizes the total
energy of the state variable $u_{\Lambda}$ to one.
\todo[inline]{What is the variable $U$? Have we defined it?}

@@

Since the TKdV equation satisfies the Liouville property, the equilibrium
invariant measure can be described by a statistical formalism
\cite{abramov2003hamiltonian,majda2006nonlinear,bajars2013weakly}
based on a Gibbs measure with the conserved energy $\mathcal{E}_{\Lambda}$ and
Hamiltonian $\mathcal{H}_{\Lambda}$. The equilibrium invariant measure is dictated
by the conservation laws in the TKdV equation. In the case of fixed energy $E_{0}$, 
this is the \emph{mixed Gibbs measure} with microcanonical energy and canonical Hamiltonian
ensembles \cite{abramov2003hamiltonian}
\begin{equation}
\mathcal{G}_{\theta}^{\pm}\left(u_{\Lambda}^{\pm};E_{0}\right)=C_{\theta}^{\pm}\exp\left(-\theta^{\pm}\mathcal{H}\left(u_{\Lambda}^{\pm}\right)\right)\delta\left(\mathcal{E}\left(u_{\Lambda}^{\pm}\right)-E_{0}\right),\label{eq:gibbs_mixed}
\end{equation}
where $\theta$ represents the ``inverse temperature''. The distinct
statistics in the upstream and downstream waves can be controlled $\theta$. 
As shown below, we find that negative temperature, $\theta^{\pm}<0$, is the appropriate regime to predict the
experiments. In the incoming wave field, $\theta^{-}$ is chosen so that $\mathcal{G}_{\theta}^{-}$
has nearly Gaussian statistics. Using the above invariant measures [\ref{eq:gibbs_mixed}],
the expectation of any functional $F\left(u\right)$ can be computed as
\[
\left\langle F\right\rangle _{\mathcal{G}_{\theta}}\equiv\int F\left(u\right)\mathcal{G}_{\theta}\left(u\right)du.
\]
The value of $\theta$ in the invariant measure is specified from
$\left\langle H_{\Lambda}\right\rangle _{\mathcal{G}_{\theta}}$ \cite{abramov2003hamiltonian,bajars2013weakly}.
\todo[inline]{I do not quite understand the above sentence.}
In addition to producing equilibrium PDFs of $u_{\Lambda}$, the invariant measure can be used to predict the equilibrium energy spectrum without the direct simulation of TKdV. Direct simulation, however, is required to recover transient statistics of $u_{\Lambda}$ and time autocorrelations.
%The invariant measure also predicts an equilibrium energy spectrum without running the TKdV equation directly. On the other hand, the time autocorrelation and transient statistics about the state variable $u_{\Lambda}$ cannot be recovered from the statistical theory.

@@

\acknow{This research of A. J. M. is partially supported by the Office of
Naval Research through MURI N00014-16-1-2161. D. Q. is supported as
a postdoctoral fellow on the second grant. M.N.J.M would like to acknowledge support from Simons grant 524259.}

>>>>

Controlled laboratory experiments were carried out in  \cite{bolles2018anomalous} to examine the statistical behavior of surface waves crossing an ADC. In these experiments, nearly unidirectional waves are generated by a paddle wheel and propagate through a long, narrow wave tank. Midway through, the waves encounter a step in the bottom topography, and thus abruptly transition to a shallower depth. The paddle wheel is forced with a pseudo-random signal intended to mimic a Gaussian random sea upstream of the ADC. In particular, the paddle angle is specified as
\[
\theta\left(t\right)=\theta_{0}+\Delta\theta\sum_{n=1}^{N}a_{n}\cos\left(\omega_{n}t+\delta_{n}\right),\;E\sim\left(\Delta\theta\right)^{2}\sum_{n}a_{n}^{2}\omega_{n}^{2}.
\]
where the weights $a_n$ are Gaussian in spectral space with peak frequency $\omega_p$ and the phases $\delta_n$ are uniformly distributed random variables. The peak frequency gives rise to a characteristic wavelength $\lambda_c$ which can be estimated from the dispersion relation. The energy $E$ injected into the system is determined by the angle amplitude $\Delta\theta$, which is the main control parameter varied in \cite{bolles2018anomalous}. Optical measurements of the free surface reveal a number of surprising statistical features:
\begin{itemize}
\item Distinct statistics are found between the incoming and outgoing wave
disturbances: incoming waves display near-Gaussian statistics,
while outgoing waves skew strongly towards positive displacement.
\item The degree of non-Gaussianity present in the outgoing waves depends on the injected energy $E$: 
larger energies generate stronger skewness in the surface displacement PDFs and more extreme events.
\item Compared to the incoming wave train, the power spectrum of the outgoing wave field decays more slowly, which indicates that the anomalous behavior is associated with an elevated level of high frequencies.
%{The waves also show different slopes in the power spectra at upstream and downstream locations}.
\end{itemize}

@@

The Korteweg-de Vries (KdV) equation is a one-dimensional, deterministic model capable of describing surface wave turbulence. More specifically, KdV is leading-order approximation for surface waves governed by a balance of nonlinear and dispersive effects, valid in an appropriate far-field limit \cite{johnson1997modern}. Moreover, KdV has been adapted to describe waves propagating over variable depth  \cite{johnson1997modern}. 
Here, we consider the variable-depth KdV equation truncated at wavenumber $\Lambda$ (with $J=2\Lambda+1$ grid points) in order to generate weakly turbulent dynamics. The surface displacement is described by the state variable $u_{\Lambda}^{\pm}\left(x,t\right)$ with superscript `$-$' for the incoming waves and `$+$' for the outgoing waves.
The Galerkin truncated variable $u_{\Lambda}=\sum_{1\leq\left|k\right|\leq\Lambda}\hat{u}_{k}\left(t\right)e^{ikx}$ is normalized with zero mean $\hat{u}_{0}=0$ and unit energy $2\pi\sum_{k=1}^{\Lambda}\left|\hat{u}_{k}\right|^{2}=1$, which are conserved quantities. Here, $u_{\Lambda}\equiv\mathcal{P}_{\Lambda}u$ denotes the subspace projection. The evolution of  $u_{\Lambda}^{\pm}$ is governed by the truncated KdV equation with depth change $D_{\pm}$
\begin{equation}
\frac{\partial u_{\Lambda}^{\pm}}{\partial t}+\frac{D_{\pm}^{-3/2}}{2}E_{0}^{1/2}L_{0}^{-3/2}\frac{\partial}{\partial x}\mathcal{P}_{\Lambda}\left(u_{\Lambda}^{\pm}\right)^{2}+D_{\pm}^{1/2}L_{0}^{-3}\frac{\partial^{3}u_{\Lambda}^{\pm}}{\partial x^{3}}=0. \label{eq:tKdV_nondim}
\end{equation}
Equation [\ref{eq:tKdV_nondim}] is non-dimensionalized on the periodic
domain $x\in\left[-\pi,\pi\right]$. The depth is assumed to be unit $D_{-}=1$ before the ADC and $D_{+}<1$ after the ADC. 
The conserved Hamiltonian can be decomposed as
\begin{align*}
\mathcal{H}_{\Lambda}^{\pm} & =  D_{\pm}^{-3/2}E_{0}^{1/2}L_{0}^{-3/2}H_{3}\left(u_{\Lambda}^{\pm}\right)-D_{\pm}^{1/2}L_{0}^{-3}H_{2}\left(u_{\Lambda}^{\pm}\right),\\
& H_{3}\left(u\right) =  \frac{1}{6}\int_{-\pi}^{\pi}u^{3}dx,\;H_{2}\left(u\right)=\frac{1}{2}\int_{-\pi}^{\pi}\left(\frac{\partial u}{\partial x}\right)^{2}dx.
\end{align*}
where we refer to $H_3$ as the cubic term and $H_2$ the quadratic term.
We introduce parameters $\left(E_{0},L_{0},\Lambda\right)$
based on the following assumptions:
\begin{itemize}
\item The wavenumber truncation $\Lambda$ is fixed at a moderate value
for generating weakly turbulent dynamics.
\item The state variable $u_{\Lambda}^{\pm}$ is normalized with zero mean
and unit energy, $\mathcal{M}\left(u_{\Lambda}\right)=0,\mathcal{E}\left(u_{\Lambda}\right)=1$,
which are conserved during evolution. Meanwhile, $E_{0}$ characterizes the total
energy injected into the system based on the driving amplitude $\Delta \theta$.
\item The length scale of the system $L_{0}$ is chosen so that the resolved scale $\Delta x=2\pi L_{0}/J$ is comparable to the the characteristic wave length $\lambda_{c}$ from the experiments.
\end{itemize}
% ORIGINAL
%The intuition for the distinct model dynamics comes from the balance
%between the cubic and quadratic terms in the Hamiltonian $\mathcal{H}_{\Lambda}^{\pm}$.
%After the depth change, $D_{+}<1$, more weight is added in the cubic
%term, $H_{3}$, for stronger nonlinearity and weaker dispersion for
%the third-order derivative term reflected by the smaller coefficient
%for $H_{2}$ in the Hamiltonian. 
Some intuition for how equation [\ref{eq:tKdV_nondim}] produces different dynamics on either side of the ADC can be gained by considering the relative contributions of $H_3$ and $H_2$ in the Hamiltonian $\mathcal{H}_{\Lambda}^{\pm}$. The depth change, $D_{+}<1$, increases the weight of $H_3$ and decreases that of $H_2$, thereby promoting the effects of nonlinearity over dispersion and creating conditions favorable for extreme events. Since $\frac{\partial u}{\partial x}$ is the slope of the wave height, $H_{2}\left(u\right)$ measures the
wave slope energy.

@@

The model parameters $\left(E_{0},L_{0},\Lambda\right)$ in [\ref{eq:tKdV_nondim}]
can be directly linked to the basic scales from the physical problem.
The important parameters that characterize the experiments \cite{bolles2018anomalous}
include: $\epsilon=\frac{a}{H_{0}}$ the wave amplitude $a$ to (upstream) water
depth $H_{0}$ ratio; $\delta=\frac{H_{0}}{\lambda_{c}}$ the water
depth to wavelength $\lambda_{c}$ ratio; and $D_{0}=\frac{d}{H_{0}}$ 
the depth ratio with upstream value $d=H_{0}$
and downstream value $d<H_{0}$. 
By comparing the characteristic physical scales, the normalized TKdV equation [\ref{eq:tKdV_nondim}] can be
linked to these experimental parameters via
\begin{equation}
L_{0}=6^{\frac{1}{3}}\left(M\epsilon^{\frac{1}{2}}\delta^{-1}\right),\;E_{0}=\frac{27}{2}\gamma^{-2}\left(M\epsilon^{\frac{1}{2}}\delta^{-1}\right),\label{eq:params}
\end{equation}
where $M$ defines the computational domain size $M\lambda_{c}$
as $M$-multiple of the characteristic wavelength $\lambda_{c}$,
and $\gamma=\frac{U}{a}$ with $U$ a scaling factor for the state variable $u_{\Lambda}$ normalizes the total energy of the system to one.

@@

Since the TKdV equation satisfies the Liouville property, the equilibrium
invariant measure can be described by a statistical formalism
\cite{abramov2003hamiltonian,majda2006nonlinear,bajars2013weakly}
based on a Gibbs measure with the conserved energy $\mathcal{E}_{\Lambda}$ and
Hamiltonian $\mathcal{H}_{\Lambda}$. The equilibrium invariant measure is dictated
by the conservation laws in the TKdV equation. In the case of fixed energy $E_{0}$, 
this is the \emph{mixed Gibbs measure} with microcanonical energy and canonical Hamiltonian
ensembles \cite{abramov2003hamiltonian}
\begin{equation}
\mathcal{G}_{\theta}^{\pm}\left(u_{\Lambda}^{\pm};E_{0}\right)=C_{\theta}^{\pm}\exp\left(-\theta^{\pm}\mathcal{H}\left(u_{\Lambda}^{\pm}\right)\right)\delta\left(\mathcal{E}\left(u_{\Lambda}^{\pm}\right)-E_{0}\right),\label{eq:gibbs_mixed}
\end{equation}
where $\theta$ represents the ``inverse temperature''. The distinct
statistics in the upstream and downstream waves can be controlled $\theta$. 
As shown below, we find that negative temperature, $\theta^{\pm}<0$, is the appropriate regime to predict the
experiments. In the incoming wave field, $\theta^{-}$ is chosen so that $\mathcal{G}_{\theta}^{-}$
has nearly Gaussian statistics. Using the above invariant measures [\ref{eq:gibbs_mixed}],
the expectation of any functional $F\left(u\right)$ can be computed as
\[
\left\langle F\right\rangle _{\mathcal{G}_{\theta}}\equiv\int F\left(u\right)\mathcal{G}_{\theta}\left(u\right)du.
\]
The value of $\theta$ in the invariant measure is specified from
$\left\langle H_{\Lambda}\right\rangle _{\mathcal{G}_{\theta}}$ \cite{abramov2003hamiltonian,bajars2013weakly}.

In addition to producing equilibrium PDFs of $u_{\Lambda}$, the invariant measure can be used to predict the equilibrium energy spectrum without the direct simulation of TKdV. Direct simulation, however, is required to recover transient statistics of $u_{\Lambda}$ and time autocorrelations.
%The invariant measure also predicts an equilibrium energy spectrum without running the TKdV equation directly. On the other hand, the time autocorrelation and transient statistics about the state variable $u_{\Lambda}$ cannot be recovered from the statistical theory.

@@

\acknow{This research of A. J. M. is partially supported by the Office of
Naval Research through MURI N00014-16-1-2161. D. Q. is supported as
a postdoctoral fellow on the grant. M.N.J.M would like to acknowledge support from Simons grant 524259.}
